---
title: "Untitled"
format: html
editor: visual
---

## First moment

$$W_{0}e^{\mu T}-e^{\mu T}\sum_{i=0}^{n}c_{i}e^{-\mu i\Delta t}$$

::: callout-note
The way the algorithm is written, we start in period 0 with $W_{0}=21$ and then immediately consume $c=1$ unit of wealth.
:::

```{r}
#| label: first moment
m1 <- function(W0, mu, dt, c) {
  n  <- length(c) - 1L          # c is c_0, c_1, ..., c_n  → n+1 elements
  T  <- n * dt
  i  <- 0:n
  exp(mu * T) * (W0 - sum(c * exp(-mu * i * dt)))
}

m1_2 <- function(W0, mu, dt, c) {
  n <- length(c) - 1L
  i <- 0:n
  W0 * exp(mu * n * dt) - sum(c * exp(mu * (n - i) * dt))
}


# correct value is ~ 64.491
m1(W0 = 21, mu = 0.07, dt = 1, c = rep(1,0,31))

m1_2(W0 = 21, mu = 0.07, dt = 1, c = rep(1,0,31))

```

## Second moment

$$ \begin{align*}
 & e^{\left(2\mu+\sigma^{2}\right)T}\left(W_{0}^{2}-2W_{0}\sum_{i=0}^{n}c_{i}e^{-\left(\mu+\sigma^{2}\right)i\Delta t}\right.\\
 & \left.+2\sum_{i=0}^{n}\sum_{j=0}^{i-1}c_{i}c_{j}e^{-\mu j\Delta t-\left(\mu+\sigma^{2}\right)i\Delta t}+\sum_{i=0}^{n}c_{i}^{2}e^{-\left(2\mu+\sigma^{2}\right)i\Delta t}\right)
\end{align*}$$

```{r}
#| label: second moment
m2 <- function(W0, mu, sigma, dt, cvec) {
  n  <- length(cvec) - 1L        # cvec = c_0, ..., c_n
  i  <- 0:n
  s2 <- sigma^2

  a <- cvec * exp(-(mu + s2) * i * dt)   # outer weights
  b <- cvec * exp(-mu * i * dt)          # inner weights
  Bprev <- c(0, head(cumsum(b), -1L))    # B_{i-1} = sum_{j<i} b_j

  exp((2 * mu + s2) * n * dt) *
    (W0^2
     - 2 * W0 * sum(a)
     + 2 * sum(a * Bprev)
     + sum(cvec^2 * exp(-(2 * mu + s2) * i * dt)))
}

# correct value is ~ 33847
m2(W0 = 21, mu = 0.07, sigma = 0.20, dt = 1, c = rep(1,0,31))
```

## Third moment

$$\begin{align*}
 & W_{0}^{3}-3W_{0}^{2}\sum_{s_{0}=0}^{n}c_{s_{0}}e^{-\left(\mu+2\sigma^{2}\right)s_{0}\Delta t}+3W_{0}\sum_{s_{1}=0}^{n}\sum_{s_{0}=0}^{n}c_{s_{1}}c_{s_{0}}e^{-\left(\mu+\sigma^{2}\right)s_{0}\Delta t-\left(\mu+2\sigma^{2}\right)s_{1}\Delta t}\\
 & -\sum_{s_{2}=0}^{n}\sum_{s_{1}=0}^{n}\sum_{s_{0}=0}^{n}c_{s_{2}}c_{s_{1}}c_{s_{0}}e^{-\mu s_{0}\Delta t-\left(\mu+\sigma^{2}\right)s_{1}\Delta t-\left(\mu+2\sigma^{2}\right)s_{2}\Delta t}
\end{align*}$$

```{r}
m3 <- function(W0, mu, sigma, dt, cvec, prefactor = 1) {
  n  <- length(cvec) - 1L        # cvec = c_0, ..., c_n
  i  <- 0:n
  s2 <- sigma^2

  A0 <- sum(cvec * exp(-(mu          ) * i * dt))
  A1 <- sum(cvec * exp(-(mu +     s2 ) * i * dt))
  A2 <- sum(cvec * exp(-(mu + 2 * s2 ) * i * dt))

  exp(3*(mu + s2 ) * n * dt) * (W0^3 - 3 * W0^2 * A2 + 3 * W0 * A1 * A2 - A0 * A1 * A2)
}

# correct value is ~ 5.82E+07 
m3(W0 = 20, mu = 0.07, sigma = 0.20, dt = 1, c = rep(1,0,31))

m3(W0 = 21, mu = log(1+0.07), sigma = 0.20, dt = 1, c = rep(1,0,31))


```

```{r}
dt = 1
sigma = 0.20
mu = 0.07
cvec = rep(1,0,31)
n  <- length(cvec) - 1L        # cvec = c_0, ..., c_n
i  <- 0:n
s2 <- sigma^2

A0 <- sum(cvec * exp(-(mu          ) * i * dt))
A1 <- sum(cvec * exp(-(mu +     s2 ) * i * dt))
A2 <- sum(cvec * exp(-(mu + 2 * s2 ) * i * dt))


sum1 <- 0
for (s_0 in 0:n){
  sum1 <- 
    sum1 + cvec[s_0+1] * exp(-(mu + 2 * s2 ) * s_0 * dt)
  
}

A2 - sum1

sum2 <- 0
for (s_1 in 0:n){
  for (s_0 in 0:n){
    sum2 <- 
      sum2 + cvec[s_1+1]*cvec[s_0+1] * exp(-(mu + s2 ) * s_0 * dt -(mu + 2 * s2 )* s_1 * dt )
  }
}

A1 * A2 - sum2

sum2 <- 0
for (s_1 in 0:n){
  for (s_0 in 0:n){
    sum2 <- 
      sum2 + cvec[s_1+1]*cvec[s_0+1] * exp(-(mu + s2 ) * s_0 * dt -(mu + 2 * s2 )* s_1 * dt )
  }
}


sum3 <- 0
for (s_2 in 0:n){
  for (s_1 in 0:n){
    for (s_0 in 0:n){
      sum3 <- 
        sum3 + 
        cvec[s_2+1]*cvec[s_1+1]*cvec[s_0+1] * 
        exp(-(mu * s_0 * dt) - (mu + s2 ) * s_1 * dt -(mu + 2 * s2 )* s_2 * dt )
    }
  }
}
  
A0 * A1 * A2 - sum3


#3147.31 * exp(3*(mu + s2 ) * n * dt)

```

```{r}
dt = 1
sigma = 0.20
mu = 0.07
cvec = rep(1,0,31)
n  <- length(cvec) - 1L        # cvec = c_0, ..., c_n
i  <- 0:n
s2 <- sigma^2
W0 <- 21

sum1 <- 0
for (s_0 in 0:n){
  sum1 <- 
    sum1 + cvec[s_0+1] * exp(-(mu + 2 * s2 ) * s_0 * dt)
  
}

sum2 <- 0
for (s_1 in 0:n){
  for (s_0 in 0:(s_1-1)){
    if(s_0 >= 0){
      sum2 <- 
        sum2 + cvec[s_1+1]*cvec[s_0+1] * exp(-(mu + s2 ) * s_0 * dt -(mu + 2 * s2 )* s_1 * dt )
    }
  }
}

sum3 <- 0
for (s_2 in 0:n){
  for (s_1 in 0:(s_2-1)){
    if(s_1 >= 0){
      for (s_0 in 0:(s_1-1)){
        if(s_0 >= 0){
          sum3 <- 
            sum3 + 
            cvec[s_2+1]*cvec[s_1+1]*cvec[s_0+1] * 
            exp(-(mu * s_0 * dt) - (mu + s2 ) * s_1 * dt -(mu + 2 * s2 )* s_2 * dt )
        }
      }
    }
  }
}

exp(3*(mu + s2 ) * n * dt) * (W0^3 - 3 * W0^2 * sum1 + 3 * W0 * sum2 - sum3)

```

```{r}
dt = 1
sigma = 0.20
mu = 0.07
cvec = rep(1,0,31)
n  <- length(cvec) - 1L        # cvec = c_0, ..., c_n
i  <- 0:n
s2 <- sigma^2
W0 <- 21

sum1 <- 0
for (s_0 in 0:n){
  sum1 <- 
    sum1 + cvec[s_0+1] * exp(-(mu + 2 * s2 ) * s_0 * dt)
  
}

sum2 <- 0
for (s_1 in 0:n){
  for (s_0 in 0:(max(s_1-1,0))){
      sum2 <- 
        sum2 + cvec[s_1+1]*cvec[s_0+1] * exp(-(mu + s2 ) * s_0 * dt -(mu + 2 * s2 )* s_1 * dt )
  }
}

sum3 <- 0
for (s_2 in 0:n){
  for (s_1 in 0:(s_2-1)){
    if(s_1 >= 0){
      for (s_0 in 0:(s_1-1)){
        if(s_0 >= 0){
          sum3 <- 
            sum3 + 
            cvec[s_2+1]*cvec[s_1+1]*cvec[s_0+1] * 
            exp(-(mu * s_0 * dt) - (mu + s2 ) * s_1 * dt -(mu + 2 * s2 )* s_2 * dt )
        }
      }
    }
  }
}

exp(3*(mu + s2 ) * n * dt) * (W0^3 - 3 * W0^2 * sum1 + 3 * W0 * sum2 - sum3)

```

$$\begin{align*}
 & W_{0}^{3}-3W_{0}^{2}\sum_{s_{0}=0}^{n}c_{s_{0}}e^{-\left(\mu+2\sigma^{2}\right)s_{0}\Delta t}+3W_{0}\sum_{s_{1}=0}^{n}\sum_{s_{0}=0}^{s_{1}-1}c_{s_{1}}c_{s_{0}}e^{-\left(\mu+\sigma^{2}\right)s_{0}\Delta t-\left(\mu+2\sigma^{2}\right)s_{1}\Delta t}\\
 & -\sum_{s_{2}=0}^{n}\sum_{s_{1}=0}^{s_{2}-1}\sum_{s_{0}=0}^{s_{1}-1}c_{s_{2}}c_{s_{1}}c_{s_{0}}e^{-\mu s_{0}\Delta t-\left(\mu+\sigma^{2}\right)s_{1}\Delta t-\left(\mu+2\sigma^{2}\right)s_{2}\Delta t}
\end{align*}$$

```{r}
excl_cumsum <- function(x) c(0, head(cumsum(x), -1L))

future_wealth_cube <- function(W0, mu, sigma, dt, cvec, prefactor = 1) {
  stopifnot(length(cvec) >= 1L, is.numeric(cvec), all(is.finite(cvec)))
  n  <- length(cvec) - 1L
  i  <- 0:n
  s2 <- sigma^2
  
  b0 <- cvec * exp(-(mu          ) * i * dt)
  b1 <- cvec * exp(-(mu +     s2 ) * i * dt)
  b2 <- cvec * exp(-(mu + 2 * s2 ) * i * dt)
  
  single <- sum(b2)                              # sum_{s0} c_{s0} e^{-(mu+2 s2) s0 dt}
  double <- sum(b2 * excl_cumsum(b1))            # sum_{s1} sum_{s0<s1}
  triple <- sum(b2 * excl_cumsum(b1 * excl_cumsum(b0)))
  
  prefactor * (W0^3 - 3 * W0^2 * single + 3 * W0 * double - triple)
}


exp(3*(mu + s2 ) * n * dt) * future_wealth_cube(W0 = 21, mu = 0.07, sigma = 0.20, dt = 1, c = rep(1,0,31))
```

## scratch

$$\begin{eqnarray*}
\left(a\alpha-b\beta\right)^{m} & = & \sum_{k=0}^{m}\left(-1\right)^{m-k}{m \choose k}a^{k}b^{m-k}\alpha^{k}\beta^{m-k}
\end{eqnarray*}$$ \## Statement

Let $a,b,\alpha,\beta$ be elements of a commutative ring (real or complex numbers being the case of interest), and let $m\in\mathbb{Z}_{\geq 0}$. Then

$$(a\alpha-b\beta)^{m}=\sum_{k=0}^{m}(-1)^{m-k}\binom{m}{k}a^{k}b^{m-k}\alpha^{k}\beta^{m-k}.$$

A preliminary observation that removes most of the apparent content: the right side depends on $a,\alpha$ only through $a^k\alpha^k$ and on $b,\beta$ only through $b^{m-k}\beta^{m-k}$. Setting $A=a\alpha$ and $B=b\beta$, the claim is exactly

$$(A-B)^m=\sum_{k=0}^m(-1)^{m-k}\binom{m}{k}A^kB^{m-k},$$

and the four-variable form is recovered from $A^k=a^k\alpha^k$, $B^{m-k}=b^{m-k}\beta^{m-k}$ — which is itself an application of commutativity. So the statement is the binomial theorem, and the only question is whether you want it derived or cited.

## Proof 1: substitution

The binomial theorem states that for commuting $x,y$ and $m\geq 0$,

$$(x+y)^m=\sum_{k=0}^m\binom{m}{k}x^ky^{m-k}.$$

Take $x=a\alpha$ and $y=-b\beta$. Then

$$(a\alpha-b\beta)^m=\sum_{k=0}^m\binom{m}{k}(a\alpha)^k(-b\beta)^{m-k}=\sum_{k=0}^m\binom{m}{k}a^k\alpha^k\,(-1)^{m-k}b^{m-k}\beta^{m-k},$$

using $(uv)^n=u^nv^n$ and $(-w)^n=(-1)^nw^n$. Rearranging the (commuting) factors gives the claim. $\blacksquare$

## Proof 2: induction on $m$, self-contained

Proof 1 defers everything to the binomial theorem, so here is the argument without that appeal. Write $A=a\alpha$, $B=b\beta$; note $AB=BA$.

**Base case** $m=0$. The left side is $1$ by convention on the empty product. The right side is the single term $k=0$: $(-1)^0\binom{0}{0}A^0B^0=1$.

**Inductive step.** Suppose $(A-B)^m=\sum_{k=0}^m(-1)^{m-k}\binom{m}{k}A^kB^{m-k}$. Then

$$(A-B)^{m+1}=A\sum_{k=0}^m(-1)^{m-k}\binom{m}{k}A^kB^{m-k}\;-\;B\sum_{k=0}^m(-1)^{m-k}\binom{m}{k}A^kB^{m-k}.$$

Pulling $A$ and $B$ inside — the step that uses $BA^k=A^kB$ —

$$=\sum_{k=0}^m(-1)^{m-k}\binom{m}{k}A^{k+1}B^{m-k}+\sum_{k=0}^m(-1)^{m-k+1}\binom{m}{k}A^{k}B^{m+1-k}.$$

Reindex the first sum by $j=k+1$ (so $j$ runs $1$ to $m+1$) and the second by $j=k$ (so $j$ runs $0$ to $m$):

$$=\sum_{j=1}^{m+1}(-1)^{m+1-j}\binom{m}{j-1}A^{j}B^{m+1-j}+\sum_{j=0}^{m}(-1)^{m+1-j}\binom{m}{j}A^{j}B^{m+1-j}.$$

Both sums now carry the same sign factor $(-1)^{m+1-j}$ and the same monomial $A^jB^{m+1-j}$, so with the standard conventions $\binom{m}{-1}=\binom{m}{m+1}=0$ (which let both sums run over $j=0,\dots,m+1$) the coefficient of $A^jB^{m+1-j}$ is

$$(-1)^{m+1-j}\left[\binom{m}{j-1}+\binom{m}{j}\right]=(-1)^{m+1-j}\binom{m+1}{j}$$

by Pascal's rule. Hence

$$(A-B)^{m+1}=\sum_{j=0}^{m+1}(-1)^{m+1-j}\binom{m+1}{j}A^{j}B^{m+1-j},$$

completing the induction. Substituting $A^j=a^j\alpha^j$ and $B^{m+1-j}=b^{m+1-j}\beta^{m+1-j}$ gives the stated form. $\blacksquare$

## What the hypotheses are actually doing

**Commutativity is the only real assumption, and it is used twice.** Once cosmetically, to split $(a\alpha)^k$ into $a^k\alpha^k$. Once essentially, in the inductive step, to move $B$ past $A^k$. If $a,b,\alpha,\beta$ were matrices or operators the identity fails at $m=2$ already: $(A-B)^2=A^2-AB-BA+B^2$, which equals $A^2-2AB+B^2$ only when $AB=BA$. Given the moment calculations in this thread the point is moot — random variables commute under multiplication — but it is worth knowing which line would break if these ever became non-commuting objects.

**The sign.** $(-1)^{m-k}=(-1)^{m+k}$ since they differ by $(-1)^{2k}$, so the identity is often written with $(-1)^{m+k}$; the two are interchangeable. The sign is attached to $b\beta$, the term carrying exponent $m-k$, which is why it appears as $(-1)^{m-k}$ rather than $(-1)^k$.

**Degenerate cases.** $m=0$ requires $x^0=1$ including at $x=0$; this is the usual convention and is what makes the $k=0$ and $k=m$ terms behave when any of $a,b,\alpha,\beta$ vanish.

I also expanded both sides symbolically for $m=0,\dots,8$ and confirmed they agree identically — not part of the proof, just a guard against a transcription slip in the statement.

If we define

$$a\alpha-b\beta=W_{0}e^{\sum_{i=0}^{n-1}\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t+\sigma\sqrt{\Delta t}\phi_{i}}-\sum_{i=0}^{n}c_{i}e^{\sum_{j=i}^{n-1}\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t+\sigma\sqrt{\Delta t}\phi_{j}}$$

and

$$\begin{eqnarray*}
a & = & W_{0}e^{\sum_{i=0}^{n-1}\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t}=W_{0}e^{\left(\mu-\frac{\sigma^{2}}{2}\right)T}\\
\alpha & = & e^{\sigma\sqrt{\Delta t}\sum_{i=0}^{n-1}\phi_{i}}=e^{\sigma\sqrt{\Delta t}S_{0}}\\
b\beta & = & \sum_{i=0}^{n}c_{i}e^{\sum_{j=i}^{n-1}\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t+\sigma\sqrt{\Delta t}\phi_{j}}\\
 & = & \sum_{i=0}^{n}c_{i}e^{\left(\mu-\frac{\sigma^{2}}{2}\right)(n-i)\Delta t+\sigma\sqrt{\Delta t}S_{i}}\\
 & \Rightarrow & b=e^{\left(\mu-\frac{\sigma^{2}}{2}\right)T}\\
 &  & \beta\equiv\sum_{i=0}^{n}c_{i}e^{-\left(\mu-\frac{\sigma^{2}}{2}\right)i\Delta t+\sigma\sqrt{\Delta t}S_{i}}
\end{eqnarray*}$$

with $S_{i}\equiv\sum_{k=i}^{n-1}\phi_{k}$ for $\phi_{k}$ IID unit Normal variables.

Thus

$$a^{k}b^{m-k}=W_{0}^{k}e^{m\left(\mu-\frac{\sigma^{2}}{2}\right)T}$$

and

$$\begin{eqnarray*}
\mathbb{E}\left[\alpha^{k}\beta^{m-k}\right] & = & \mathbb{E}\left[e^{k\sigma\sqrt{\Delta t}S_{0}}\times\prod_{i=k+1}^{m}\beta_{i}\right]\\
 &  & \mathbb{E}\left[\sum_{\kappa_{m}=0}^{n}\cdots\sum_{\kappa_{k+1}=0}^{n}\left(\prod_{i=k+1}^{m}c_{\kappa_{i}}\right)e^{-\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t\sum_{i=k+1}^{m}\kappa_{i}+\sigma\sqrt{\Delta t}\left(kS_{0}+\sum_{i=k+1}^{m}S_{\kappa_{i}}\right)}\right]
\end{eqnarray*}$$

and we note that the summand is a symmetric function

$$\varphi\left(\kappa_{k+1},\ldots,\kappa_{m}\right)=\left(\prod_{i=k+1}^{m}c_{\kappa_{i}}\right)e^{-\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t\sum_{i=k+1}^{m}\kappa_{i}+\sigma\sqrt{\Delta t}\left(kS_{0}+\sum_{i=k+1}^{m}S_{\kappa_{i}}\right)}$$

and so, wlog we can take $\kappa_{k+1}\le\ldots\le\kappa_{m}$. Now while the sums implicit in the $S_{\kappa_{i}}$ terms on the right are overlapping as written, we can re-arrange them into non-overlapping sums, in which form the exponential is a product of independent random variables

$$\begin{eqnarray*}
\varphi\left(\kappa_{k+1},\ldots,\kappa_{m}\right) & = & \left(\prod_{i=k+1}^{m}c_{\kappa_{i}}\right)e^{-\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t\sum_{i=k+1}^{m}\kappa_{i}+\sigma\sqrt{\Delta t}\left(kS_{0}+\sum_{i=k+1}^{m}S_{\kappa_{i}}\right)}\\
 & = & \left(\prod_{i=k+1}^{m}c_{\kappa_{i}}\right)e^{-\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t\sum_{i=k+1}^{m}\kappa_{i}+\sigma\sqrt{\Delta t}\left(k\left(S_{0}-S_{\kappa_{k+1}}\right)+\sum_{i=k+1}^{m}i\left(S_{\kappa_{i}}-S_{\kappa_{i+1}}\right)\right)}
\end{eqnarray*}$$

and $S_{\kappa_{i}}-S_{\kappa_{i+1}}$ represents the sum of $\kappa_{i+1}-\kappa_{i}$ IID unit normals, with $\kappa_{m+1}\equiv n$. Now we can take the expected value

$$\mathbb{E}\left[\varphi\left(\kappa_{k+1},\ldots,\kappa_{m}\right)\right]=\left(\prod_{i=k+1}^{m}c_{\kappa_{i}}\right)e^{-\left(\mu-\frac{\sigma^{2}}{2}\right)\Delta t\sum_{i=k+1}^{m}\kappa_{i}+\frac{\sigma^{2}}{2}\Delta t\left(k^{2}\kappa_{k+1}+\sum_{i=k+1}^{m}i^{2}\left(\kappa_{i+1}-\kappa_{i}\right)\right)}$$

and re-arranging the terms

$$\begin{eqnarray*}
  &  & \frac{\sigma^{2}}{2}\Delta t\sum_{i=k+1}^{m}\kappa_{i}+\frac{\sigma^{2}}{2}\Delta t\left(k^{2}\kappa_{k+1}+\sum_{i=k+1}^{m}i^{2}\left(\kappa_{i+1}-\kappa_{i}\right)\right)\\
  & = & \frac{\sigma^{2}}{2}\Delta t\left(k^{2}\kappa_{k+1}+\sum_{i=k+1}^{m}\kappa_{i}+i^{2}\left(\kappa_{i+1}-\kappa_{i}\right)\right)\\
  & = & -\frac{\sigma^{2}}{2}\Delta t\left(-k^{2}\kappa_{k+1}+\sum_{i=k+1}^{m}i^{2}\left(\kappa_{i}-\kappa_{i+1}\right)-\kappa_{i}\right)\\
 & = & -\frac{\sigma^{2}}{2}\Delta t\left(\sum_{i=k+1}^{m}\left(i^{2}-\left(i-1\right)^{2}-1\right)\kappa_{i}-m^{2}n\right)\\
 & = & -\left(\sum_{i=k+1}^{m}\left(i-1\right)\kappa_{i}\right)\sigma^{2}\Delta t+m^{2}\frac{\sigma^{2}}{2}T
 \end{eqnarray*}
 $$ since

$$i^{2}-\left(i-1\right)^{2}-1=i^{2}-i^{2}+2i-1-1=2\left(i-1\right)$$ so the last term can be brought out with the $a^{k}b^{m-k}=W_{0}^{k}e^{m\left(\mu-\frac{\sigma^{2}}{2}\right)T}$ term to give $$W_{0}^{k}e^{m\left(\mu+\left(m-1\right)\frac{\sigma^{2}}{2}\right)T}$$ which multiplies $$\sum_{\kappa_{m}=0}^{n}\cdots\sum_{\kappa_{k+1}=0}^{n}\left(\prod_{i=k+1}^{m}c_{\kappa_{i}}\right)e^{-\mu\Delta t\left(\sum_{i=k+1}^{m}\kappa_{i}\right)-\sigma^{2}\Delta t\left(\sum_{i=k+1}^{m}\left(i-1\right)\kappa_{i}\right)}$$
